---
title: "Chapter 0: Preliminaries"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chapter 0: Preliminaries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE}
library(BEDCA)
```

## Introduction

This document describes the use of the statistical application R as computing support in an introductory statistics course based on the text *Basic Engineering Data Collection and Analysis*. We demonstrate how R can be used to reproduce the results from many of the examples in the text.

A desirable feature of this text is the number of examples and exercises based on real data sets. Most of these are drawn from actual experiments performed by students as course projects. Some data sets are large and have a complex structure. Although it is not difficult to enter these data into a computer package like R, the process is tedious, error-prone, and provides a poor learning experience. We have provided copies of the data sets for most exercises and examples in a package named `BEDCA` that can be used with R. This document is also part of the `BEDCA` package.

Data for exercises were constructed by importing the supplied tab-delimited text files into R. In most cases the data were imported "as is," so they may require additional processing before analysis. For the most part this involves reshaping into tidy format. Data for most examples appear as tables in the body of the text; these were copied or transcribed. Data sets are identified as shown in the following examples:

-   `ch1_xmp07.RData`: Chapter 1, Example 7 (p. 16)

-   `ch3_sec1_prob1.RData`: Section 3.1, Exercise 1 (p. 77)

-   `ch3_eoc_prob01.RData`: Chapter 3, Exercise 1 (p. 113)

-   `table3_04.RData`: Table 3.4 (p. 74)

-   `table_a_2`: Appendix A, Table 2 (p. 768)

Since the text is in final form, future revisions to this package will consist of updates to these vignettes.

This document assumes that you have both R and RStudio installed on your system. If not, go to <https://posit.co/download/rstudio-desktop/> and follow the instructions there. Once you have both applications installed, launch RStudio and make the following changes:

Go to **Tools \> Global Options...** On the **General** page, go to **Save workspace to .RData** **on exit** and select **Never**.

Click **Code** on the left side, then the **Editing** tab. Check **Use native pipe operator**.

While in **Code** select the **Display** tab. Check **Show margin**. In the Editor pane you'll see a vertical line at 80 columns. Keeping your code within 80 columns makes code easier to read when printed to paper.

I recommend having the Editor and Console panes side-by-side. Select **View \> Panes \> Console on Right**.

You may also want to adjust settings to your own taste. Go to **Tools \> Global Options...** and select **Appearance**. Among other things, you can select an editor theme. This is useful if you prefer a dark background. Select a theme you like.

Besides this package, you will need to install other packages. In the **Console** window, you'll see a **\>** prompt following the R startup message. Click to the right of the prompt and you'll see a blinking cursor. Copy the following, paste it at the cursor, and hit **Enter**.

```{r, eval=FALSE}
install.packages(c("BEDCA", "broom", "broom.helpers", "DoE.base", "emmeans",
                   "fastDummies", "FrF2", "GGally", "ggDoE","ggfortify", "here", 
                   "infer", "mosaic", "multcomp", "openintro", "openxlsx", 
                   "readxl", "scales", "tidyverse", "tinytex", "writexl"))
```

Package installation needs to be done only once. At this point you're ready to do analysis. The first step is to load the necessary packages with the `library()` function; for example:

```{r, eval=FALSE}
library(BEDCA)
library(mosaic)
library(tidyverse)
```

Readers should be familiar with the following basic R concepts:

-   Data types

    -   Numeric and integer
    -   Character
    -   Logical

-   Data structures

    -   Vectors
    -   Data frames
    -   Tibbles

-   Objects and assignment

-   Functions and arguments

-   Packages, installing and loading

-   Factors and levels

-   Using the pipe |> for sequential operations

Readers not familiar with these should consult one of the many online free tutorials such as [R Basics: Quick and Easy](https://www.sthda.com/english/wiki/r-basics-quick-and-easy) from STHDA.com. Within R, help on functions or data sets is obtained with `help("name")` or `?name`.

Most analyses are done with the `tidyverse` suite of packages and the `mosaic` package. The `tidyverse` packages most often used are

-   `dplyr` for summarizing and transforming data

-   `tidyr` for reshaping

-   `ggplot2` for graphing

The `mosaic` and `ggformula` packages provide a formula interface to common statistical analyses and `ggplot` graphics. The formula interface provides a consistent approach to statistical analyses.

Data sets appearing in multiple exercises are found under first appearance; for example, data for Chapter 9, Exercise 10 are found under Chapter 4, Exercise 24.

For details of analyses, see the other vignettes.

## Bibliography

Nicholas J. Horton, Randall Pruim, and Daniel T. Kaplan (2018): [A Student's Guide to R](https://github.com/ProjectMOSAIC/LittleBooks/raw/master/StudentGuide/MOSAIC-StudentGuide.pdf). Edition 1.3.

Daniel T. Kaplan and Randall Pruim, [Statistical Modeling: A Fresh Approach](https://statistical-modeling.netlify.app)

Harvey Motulsky: *Intuitive Biostatistics: A Nonmathematical Guide to Statistical Thinking* (4th. ed.) Oxford University Press. ISBN 978-0-19-064356-0 

Randall Pruim, Daniel T. Kaplan, and Nicholas J. Horton (2017): "The `mosaic` package: Helping students to think with data using R", *The R Journal* 9(1), 77-102, <https://doi.org/10.32614/RJ-2017-024>

Stephen B. Vardeman and J. Marcus Jobe (2001), *Basic Engineering Data Collection and Analysis*, Pacific Grove, CA: Brooks/Cole. Republished 2023 by [Iowa State University Digital Press](https://www.iastatedigitalpress.com/plugins/books/127/). ISBN 978-1-958291-03-0

Stephen B. Vardeman and J. Marcus Jobe (2016), *Statistical Methods for Quality Assurance* (2nd. ed.), New York: Springer-Verlag. ISBN 978-0-387-79105-0

Hadley Wickham, Mine Çetinkaya-Rundel, and Garrett Grolemund, [R for Data Science: Import, Tidy, Transform, Visualize, and Model Data](https://r4ds.hadley.nz)
(2nd. ed.), Santa Rosa, CA: O'Reilly. ISBN 978-1-4920-9740-2
