---
title: "Chapter 0: Preliminaries"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ch0-prelim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(BEDCA)
```

## Chapter 0: Preliminaries

This document describes the use of the statistical application R as computing support in an introductory statistics course based on the text *Basic Engineering Data Collection and Analysis* by Stephen B. Vardeman and J. Marcus Jobe (Brooks/Cole 2001, republished 2023 by [Iowa State University Digital Press](https://www.iastatedigitalpress.com/plugins/books/127/)). We demonstrate how R can be used to reproduce the results in many of the examples in the text.

A desirable feature of this text is the number of examples and exercises based on real data sets. Most of these are drawn from actual experiments performed by students as course projects. As they are real data, some of the data sets are large and have a complex structure. Although it is not dicult to enter these data into a computer package like R, the process is tedious and error-prone. Furthermore, it is not much of a learning experience. We have provided copies of the data sets for the exercices and most examples in a \package", named `BEDCA`, that can be used with R. This document is also part of the \`BEDCA\` package.
