---
title: "Chapter 3: Elementary Descriptive Statistics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chapter 3: Elementary Descriptive Statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center"
  )
```

```{r setup}
library(BEDCA)
```

Chapter 3 discusses elementary graphical and numerical summaries: dotplots, histograms, run charts, quantiles, mean, standard deviation. Stem-and-leaf plots are omitted from this vignette. We also present a modeling approach to obtaining summary statistics.

Example 1 (p. 67) of this chapter revisits the heat-treating process for gears introduced in Chapter 1. As noted there, the data for the two gear treatments appear in separate columns, which is not amenable for analysis. Data must be reshaped into tidy format, with one column (`Treatment`) for the treatment type and a second (`Runout`) for the runout measurements.

```{r}
gears <- table1_01 |> 
    tidyr::pivot_longer(cols = everything(), 
                        values_to = "Runout", 
                        names_to = "Treatment") |> 
  dplyr::filter(!is.na(Runout))
mosaic::dotPlot(~ Runout | Treatment, data = gears)
```

Example 2 discusses a study to compare penetration depths of .45 caliber bullets fired into oak wood. Data appear in Table 3.1 (p. 67) and Chapter 3, Section 1, Exercise 2 (p. 77).

```{r}
bullets <- ch3_sec1_prob2 |> 
      tidyr::pivot_longer(cols = everything(), 
                        values_to = "depth", 
                        names_to = "type")
mosaic::dotPlot(~ depth | type, data = bullets,
                nint = 12)
```

